---
title: "Aggregating Data"
author: "Gaston Sanchez"
output: html_document
---

A very common task when analyzing data is data aggregation. In other words, summary statistics obtained conditional to some group or variable.

```{r}
head(iris)
```

There are three species of iris

```{r}
levels(iris$Species)
```


## Average `Sepal.Length` by species

Following a traditional programming approach, we can use a `for` loop to obtain the average sepal length of each species:
```{r}
# vector of unique species
species <- levels(iris$Species)

# average sepal length (empty vector)
asl <- numeric(length(species))

for (s in 1:length(species)) {
  aux <- iris$Species == species[s]
  asl[s] <- mean(iris$Sepal.Length[aux])
}
names(asl) <- species
asl
```

R, however, provides other alternatives to compute the average of sepal length. One option is to use `tapply()`

```{r}
tapply(iris$Sepal.Length, iris$Species, mean)
```

Instead of applying the `mean()` function, we could apply `summary()`
```{r}
tapply(iris$Sepal.Length, iris$Species, summary)
```

The same summary, but in different output format, can be obtained with `aggregate()`:
```{r}
aggregate(iris$Sepal.Length, list(iris$Species), summary)
```

Here's another example:
```{r}
tapply(iris$Sepal.Length, iris$Species, which.max)

aggregate(iris$Sepal.Length, list(iris$Species), which.max)
```

-----

Load the following packages:

```{r}
library(readr)
library(stringr)
```

```{r}
# import star-wars data
sw <- read_csv('https://raw.githubusercontent.com/gastonstat/stat133/master/datasets/starwars.csv')
```

```{r}

aggregate(sw[ ,c('height','weight')], list(sw$species), mean)
```

```{r}
aggregate(sw[ ,c('height','weight')], 
          list(sw$gender, sw$jedi), mean, na.rm=TRUE)
```

